Some forms of uncertainty may suppress the evolution of social learning

### 2022-04-13 ms.tex
outlined analysis section

% \documentclass[11pt,letterpaper]{article}
\documentclass[letterpaper,11.5pt]{scrartcl}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage{amssymb,amsmath}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
% \usepackage[margin=1.00in]{geometry}
\usepackage[margin=0.90in]{geometry}

@@

@@

We operationalized uncertainty in four different ways: 
(1) payoff ambiguity, $A_\pi$, which measures the difference
between the optimal expected payoff behavior $\pi_{high}$ and the expected payoff
of the other behaviors, $\pi_{low}$; 
(2) environmental variability, $u$, the probability the optimal behavior changes from one generation to another; 
(3) the number of possible behaviors, $B$; and 
(4) the lifespan, or time steps per generation, $M$.

@@

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agents track the mean payoff of each behavior $b$, denoted $\bar\pi_b$, and a \emph{count} of how many times it has performed each behavior, denoted $c_b$. 
Agents' beliefs about mean payoffs are updated from $\bar\pi_b$ to $\bar\pi_b'$ using exponential weighted
averaging, $\bar\pi_b' = \bar\pi_b + \frac{\mathrm{Bandit_b(0, 1)} - \bar\pi_b}{c_b'}$, where
$\mathrm{Bandit_b(0, 1)}$ is 0 or 1 depending on the result of the bandit draw for behavior $b$. 
$\bar\pi_b$ is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_b$ is initialized as that of its teacher selected
through performance-biased oblique learning.

@@

In our model uncertainty is a tunable consequence of four environmental features:
%that structure the behavior of agents.  
% Many factors contribute to the 
% uncertainty involved in both social and asocial learning. 
%We model uncertainty in four distinct ways: 
\begin{enumerate}
    \item We vary the frequency at which the optimal behavior changes, which we call \emph{environmental variability}, $u$. At the start of each generation, with probability $u$, a new behavior is assigned a payoff of $\pi_{high}$ while all other behaviors are assigned a payoff of $\pi_{low}$. Otherwise, the same behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pi_{high}$, while all other behaviors yield a 
      payoff of $\pi_{low} < \pi_{high}$. 
      The difference between these quantities is the \emph{payoff ambiguity}, 
    \begin{equation}
      A_\pi = 1 - (\pi_{high} - \pi_{low}),
      \label{eq:payoffAmbiguity}
    \end{equation}
    which is maximized when the expected payoffs are close to equal and minimized
    when they differ greatly. For simplicity we will set $\pi_{high} = 0.9$
    for all analyses and vary $\pi_{low}$ in order to vary $A_\pi$.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  %risk, as it becomes increasingly costly to experiment with alternatives. \cm{is this risk if it reduces variance of outcomes? it sounds like it's just bad for there to be more 0 payoff options}
    \item Finally, we vary number of behavioral events per generation, $M$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($M$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

@@

\begin{figure}
  \centering
    \includegraphics[width=0.95\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
  \caption{Model dynamics are composed of two major parts: first, agents perform
  behaviors for $M$ time steps in a single generation. Then agents reproduce
  randomly, with more successful agents being more likely to reproduce. Newly
  ``born'' social learner agents use success bias to select and learn from a chosen
  teacher. Then, all agents from the previous generation die off, all new agents
  begin a new generation, and the process continues until 1000 total time steps
  have elapsed. At model initialization all agents are ``blank slates''.}
  \label{fig:IntraInterGenerationalDynamics}
\end{figure}

@@

We manipulated environmental uncertainty parameters described above, $u$,
$\pi_{low}$, $B$, and $M$, to examine their effects on our
main outcome measure, the prevalence of social learning $s$. For each parameter setting
in our analysis we calculated the average value of $s$ at the final time step
of $T=1000$ across 100 runs for each combination of uncertainty parameter values.
The model

@@

To analyze the effect of various forms of uncertainty on social learning 
prevalence, we analyzed the outcomes of several model runs initialized with
different uncertainty parameters $u$, $A_\pi$ (via $\pi_{low}$), 
and $M$ (AND SOON $B$???). Specifically, for each parameter setting over
$\pi_{low} \in \{0.1, 0.5, 0.8\}$ and $M \in \{1, 2, 4\}$, we varied 
$u \in \{0.0, 0.2, \ldots, 1.0\}$.
Our analysis confirmed existing theoretical predictions that social learning
is valuable only when environmental uncertainty is sufficiently small. However
our analysis also revealed that environmental uncertainty is not the only 
relevant factor. Payoff ambiguity can alter selection selection pressure for or
against social learning, depending on the number of time steps per generation.
We tested three values of $\pi_{low} \in \{0.1, 0.5, 0.8\}$, representing
low, moderate, and high values of $A_\pi$.

When $\pi_{low} = 0.1$, $M$ had little to no effect (Figure~\ref{fig:results_pilow_0p1}).
For all three values of $M$, social learning decreases from 1 dramatically over moderate
values of $u$, going to 0 by $u=0.6$. When $\pi_{low}=0.5$, however, the 
transition from $\langle s \rangle = 1$ to $\langle s \rangle = 0$ is not as
sharp, and it is slower for $M=1$ compared to $M=2$ and $M=4$. When $\pi_{low} = 0.8$
there appears to be a weak dependence of social learning on $u$ across $M$,
but this trend is mostly dominated by noise.
(EXPAND THIS---PROBABLY NEED B > 2, WHAT ELSE?)

(EXPLAIN WHY)

@@

% \bibliography{/Users/mt/workspace/Writing/library.bib}
\bibliography{this.bib}

>>>>

% \documentclass[11pt,letterpaper]{article}
\documentclass[letterpaper,11.5pt]{scrartcl}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage{amssymb,amsmath}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
\usepackage[margin=1.00in]{geometry}
% \usepackage[margin=0.90in]{geometry}

@@

\newcommand{\pisub}[1]{\pi_{\mathrm{#1}}}
\newcommand{\pilow}{\pisub{low}}

\newcommand{\meansl}{\langle s \rangle}

@@

We operationalized uncertainty in four different ways: 
(1) payoff ambiguity, $A_\pi$, which measures the difference
between the optimal expected payoff behavior $\pisub{high}$ and the expected payoff
of the other behaviors, $\pisub{low}$; 
(2) environmental variability, $u$, the probability the optimal behavior changes from one generation to another; 
(3) the number of possible behaviors, $B$; and 
(4) the lifespan, or time steps per generation, $M$.

@@

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agents track the mean payoff of each behavior $b$, denoted $\bar\pi_b$, and a \emph{count} of how many times it has performed each behavior, denoted $c_b$. 
Agents' beliefs about mean payoffs are updated from $\bar\pi_b$ to $\bar\pi_b'$ using exponential weighted
averaging, $\bar\pi_b' = \bar\pi_b + \frac{\mathrm{Bandit_b(0, 1)} - \bar\pi_b}{c_b'}$, where
$\mathrm{Bandit}_b(0, 1)$ is 0 or 1 depending on the result of the bandit draw for behavior $b$. 
$\bar\pi_b$ is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_b$ is initialized as that of its teacher selected
through performance-biased oblique learning.

@@

In our model uncertainty is a tunable consequence of four environmental features:
%that structure the behavior of agents.  
% Many factors contribute to the 
% uncertainty involved in both social and asocial learning. 
%We model uncertainty in four distinct ways: 
\begin{enumerate}
    \item We vary the frequency at which the optimal behavior changes, which we call \emph{environmental variability}, $u$. At the start of each generation, with probability $u$, a new behavior is assigned a payoff of $\pisub{high}$ while all other behaviors are assigned a payoff of $\pisub{low}$. Otherwise, the same behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pisub{high}$, while all other behaviors yield a 
      payoff of $\pisub{low} < \pisub{high}$. 
      The difference between these quantities is the \emph{payoff ambiguity}, 
    \begin{equation}
      A_\pi = 1 - (\pisub{high} - \pisub{low}),
      \label{eq:payoffAmbiguity}
    \end{equation}
    which is maximized when the expected payoffs are close to equal and minimized
    when they differ greatly. For simplicity we will set $\pisub{high} = 0.9$
    for all analyses and vary $\pisub{low}$ in order to vary $A_\pi$.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  %risk, as it becomes increasingly costly to experiment with alternatives. \cm{is this risk if it reduces variance of outcomes? it sounds like it's just bad for there to be more 0 payoff options}
    \item Finally, we vary number of behavioral events per generation, $M$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($M$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

@@

\begin{figure}
  \caption{Model dynamics are composed of two major parts: first, agents perform
  behaviors for $M$ time steps in a single generation. Then agents reproduce
  randomly, with more successful agents being more likely to reproduce. Newly
  ``born'' social learner agents use success bias to select and learn from a chosen
  teacher. Then, all agents from the previous generation die off, all new agents
  begin a new generation, and the process continues until 1000 total time steps
  have elapsed. Agents are initialized as ``blank slates''.}
  \centering
    \includegraphics[width=0.9\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
  \label{fig:IntraInterGenerationalDynamics}
\end{figure}

@@

We manipulated environmental uncertainty parameters described above, $u$,
$\pisub{low}$, $B$, and $M$, to examine their effects on our
main outcome measure, the prevalence of social learning $s$. For each parameter setting
in our analysis we calculated the average value of $s$ at the final time step
of $T=1000$ across 100 runs for each combination of uncertainty parameter values.

To analyze the effect of various forms of uncertainty on social learning 
prevalence in our Analysis, we initialized several model runs with
different uncertainty parameters $u$, $\pisub{\mathrm{low}}$, $B$,
and $M$. We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$ for each combination of
the following parameters:
$\pisub{low} \in \{0.1, 0.5, 0.8\}$, $B \in \{2, 4, 10\}$, $M \in \{B/2, B, 2B\}$.

@@

\begin{itemize}
  \item 
    To understand how different forms of uncertainty affect social learning evolution
    we systematically varied environmental variables in our model and observed
    corresponding changes in mean social learning prevalence, $\meansl$. We
    confirmed that, in many cases, increased environmental variability $u$ 
    suppresses social learning. However\ldots

  \item 
    Our analysis confirmed existing theoretical predictions that social learning
    is valuable only when environmental variability $u$ is sufficiently small.
    Because of the central role of $u$, it is set on the x-axis in
    all plots shown in Figure~\ref{fig:results}. 
    The threshold at which social learning stops being strongly selected for
    varies depending on the value of other environmental parameters.

  \item 
    However, more possible behaviors $B$ ($B=2,4,10$ for the first, second, and
    third column respectively in Figure~\ref{fig:results}) can lead
    social learning to be selected for even for large values of $u$, possibly
    due to the low risk and potentially high reward of learning an accurate
    tally of mean payoffs from the previous generation. 
    Since there are so many behaviors and so
    few observations, softmax weighting will not significantly 
    favor a sub-optimal behavior. However, if one does learn an accurate tally
    then one does have an increased chance of performing the best one.
    \begin{itemize}
      \item 
        (THIS MAKES SENSE, BUT THE EFFECT IS TOO STRONG. WHY SHOULD THERE BE ANY
        SOCIAL LEARNING WHEN U = 1.0?  I THINK SOMETHING IS SLIGHTLY WRONG WITH THE
        MODEL---I'M WORKING ON CHECKING WHAT THAT MIGHT BE)
      \item
        ONE THING TO CHECK IS THE FRACTION DOING THE OPTIMAL BEHAVIOR WHEN
        $u=1$. IF MANY START OUT DOING OPTIMAL, THEN SOMETHING MAY BE WRONG.
    \end{itemize}

  \item 
    $\pilow$ can alter selection selection pressure for or
    against social learning, depending on the number of time steps per generation.
    For instance, when $B=2$, increasing $\pilow$ (i.e. increased uncertainty
    about which behavior is optimal) leads at first to less sharp sigmoid curve, 
    where social learning is less prevalent with increased $\pilow$ when
    $u < 0.5$ and social learning is more prevalent with increased $\pilow$
    when $u > 0.5$. When $\pilow = 0.8$ noise and genetic drift dominate across
    tested values of $B$, with possibly a slight inverse correlation 
    in social learning as $u$ increases. When $B \in \{4,10\}$, increasing
    $\pilow \in \{0.1, 0.5\}$ leads uniformly to increased $\meansl$ over 
    $u$ in those settings.

  \item
    If $M/B$ has an effect it is mostly to 
    increase reliance on social learning due to social information being of 
    higher quality since teachers would have had more learning opportunities
    for larger $M$. (THERE ARE CASES WHERE THIS IS NOT THE CASE BUT THERE IS 
    STILL TOO MUCH NOISE DUE TO SMALL NTRIALS)

  \item
    Synthesis/analysis conclusion

\end{itemize}

@@

\bibliography{/Users/mt/workspace/Writing/library.bib}
% \bibliography{this.bib}
